---
title: "Project Data"
author: "Kathryn Bernard"
date: '2022-07-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
### Load packages
```{r, message=FALSE}
library(httr) # for pulling the Strava data
library(rStrava) # for formatting the Strava data
library(tidyverse)
```

## Strava Data
### Pull data from Strava
``` {r}
# Strava app setup
# Run once, then comment out (it doesn't work in the knitted Rmd file because it requires authentication
# in a browser window to finish pulling the data from the Strava website)
#client_id <- 80261
#secret <- "****" # fill in as needed
#app_name <- "Training_Analysis"
# launches a browser window where you have to manually authenticate the app
#stoken <- httr::config(token = strava_oauth(app_name, client_id, secret, app_scope="activity:read_all"))

# extract data
#my_acts <- get_activity_list(stoken)
#df <- compile_activities(my_acts)

# save as a csv to read in the knitted version of the file
#write.csv(df,"strava.csv") 

df <- read.csv("strava.csv")

# drop unneeded columns
drop <- names(df) %in% c("athlete.id", "athlete.resource_state", "commute",
                         "display_hide_heartrate_option", "flagged", "from_accepted_tag", "has_heartrate", "has_kudoed",
                         "heartrate_opt_out", "manual", "map.id", "map.resource_state",
                         "photo_count", "private", "resource_state",
                         "start_date_local", "timezone", "total_photo_count", "trainer",
                         "utc_offset", "visibility", "average_watts", "device_watts",
                         "end_latlng1", "end_latlng2", "external_id", "gear_id", "kilojoules",
                         "map.summary_polyline", "start_latlng1", "start_latlng2", "upload_id", "upload_id_str",
                         "type", "workout_type"
                         )
df <- df[!drop]

# filter to only running activities
runs <- df %>% filter(sport_type=="Run")

# get number of observations
nrow(runs)
```

### Convert units and datatypes as needed
```{r, warning=FALSE}
# adjust datatypes to numeric 
runs <- runs %>% mutate_at(c("average_cadence", "average_heartrate", "average_temp", "average_speed", "max_heartrate", "pr_count", "suffer_score"), as.numeric, na.rm=TRUE)

# convert units
runs <- runs %>% mutate(distance_miles=distance/1.609, # distance is originally in km; convert to miles
                        temp_f=average_temp*9/5 +32, # temp is originally in C; convert to F
                        elapsed_time_min=elapsed_time/60, # time is originally in seconds; convert to minutes
                        moving_time_min=moving_time/60,
                        average_speed_mpm=round(96.56064/average_speed,2), # speed is in km/hour, we want miles/min
                        max_speed_mpm=round(96.56064/max_speed,2),
                        total_elev_gain_ft=total_elevation_gain*3.281, # elevation is in meters, we want feet
                        elev_high_ft=elev_high*3.281,
                        elev_low_ft=elev_low*3.281
                        )
# split out minutes and hours into a more typical running pace format, for readability
runs <- runs %>% separate(average_speed_mpm, into=c("average_speed_mins", "average_speed_secs"), remove=FALSE)
runs <- runs %>% separate(max_speed_mpm, into=c("max_speed_mins", "max_speed_secs"), remove=FALSE)
runs <- runs %>% mutate(average_speed_mins=as.numeric(average_speed_mins),
                        max_speed_mins=as.numeric(max_speed_mins),
                        max_speed_secs=as.numeric(formatC(round(as.numeric(max_speed_secs)/100*60,2),width=2,format="d",flag="0")),
                        average_speed_secs=as.numeric(formatC(round(as.numeric(average_speed_secs)/100*60,2),width=2,format="d",flag="0")))
```

### Summarize variables of interest
```{r, warning=FALSE}
# define a subset of variables that will potentially be used in the final analysis
summary <- c("achievement_count", "athlete_count", "average_cadence", "average_heartrate", "max_heartrate", "suffer_score",
             "total_elev_gain_ft", "distance_miles", "temp_f", "elapsed_time_min", "moving_time_min", "average_speed_mpm", "max_speed_mpm")

# get summary statistics for each
summary(runs[,summary])
```


## Garmin Data
### Read in Garmin data (downloaded manually from Garmin connect)
Note - I'm still waiting for a full data export from Garmin, so this is just a prototype based on weekly aggregated data that's available for immediate download. The final data will be different.
```{r, warning=FALSE, message=FALSE}
calories <- read.csv("Garmin_Calories.csv")
rhr <- read.csv("Garmin_RestingHeartRate.csv")
sleep <- read.csv("Garmin_Sleep.csv")
stress <- read.csv("Garmin_Stress.csv")

# combine into one dataframe based on date
calories <- calories %>% filter(!row_number()==7) # remove Aug 9 (only present in calories and not other dataframes)

require(plyr)
garmin <- join_all(list(calories,rhr,sleep,stress),by="X",type="full") # create one larger dataframe joined on column X (date)
```

### Convert units and datatypes as needed
```{r, warning=FALSE}
# sleep column is a character, want to split out into numbers for hours/minutes
garmin <- garmin %>% separate(col=Sleep, into=c("Sleep.Hours","Sleep.Minutes")) # this gives a warning about discarding missing pieces but it's fine 
# the piece that is discarded is the "hrs" unit that's added to the end of each measurement, which we don't want anyway

# rename total calories column to be more clear
garmin <- garmin %>% dplyr::rename(Total.Calories=Total)

garmin <- garmin %>% mutate(Sleep.Hours=as.numeric(Sleep.Hours),
                            Sleep.Minutes=as.numeric(Sleep.Minutes),
                            Total.Sleep=round(((Sleep.Hours*60)+Sleep.Minutes)/60,2))
```

### Summarize variables of interest
```{r, warning=FALSE}
# define a subset of variables that will potentially be used in the final analysis
summary <- c("Active.Calories", "Resting.Calories", "Total.Calories", "Resting.Heart.Rate", "Stress", "Total.Sleep")
summary(garmin[,summary])
```


 