---
title: "Project Data"
author: "Kathryn Bernard"
date: '2022-07-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
The goal of this project is to create a multiple linear regression model exploring the effects of several variables on running pace. These variables are: distance, temperature, heartrate during the run, hours of sleep the night before, and Strava's calculated "suffer score" for the activity (essentially a perceived exertion score).

This project also includes descriptive statistics for all the variables of interest, and a deeper exploration into the temperature variable.

## Setup
### Load packages
```{r, message=FALSE}
library(tidyverse)
library(RColorBrewer)
```

## Strava Data
``` {r}
strava <- read.csv("data/strava.csv")

# drop unneeded columns
drop <- names(strava) %in% c("athlete.id", "athlete.resource_state", "commute",
                         "display_hide_heartrate_option", "flagged", "from_accepted_tag", "has_heartrate", "has_kudoed",
                         "heartrate_opt_out", "manual", "map.id", "map.resource_state",
                         "photo_count", "private", "resource_state",
                         "start_date_local", "timezone", "total_photo_count", "trainer",
                         "utc_offset", "visibility", "average_watts", "device_watts",
                         "end_latlng1", "end_latlng2", "external_id", "gear_id", "kilojoules",
                         "map.summary_polyline", "start_latlng1", "start_latlng2", "upload_id", "upload_id_str",
                         "type", "workout_type"
                         )
strava <- strava[!drop]

# filter to only running activities
runs <- strava %>% filter(sport_type=="Run")

# get number of observations
nrow(runs)
```

### Convert units and datatypes as needed
```{r, warning=FALSE}
# adjust datatypes to numeric 
runs <- runs %>% mutate_at(c("average_cadence", "average_heartrate", "average_temp", "average_speed", "max_heartrate", "pr_count", "suffer_score"), as.numeric, na.rm=TRUE)

# convert units
runs <- runs %>% mutate(distance_miles=distance/1.609, # distance is originally in km; convert to miles
                        temp_f=average_temp*9/5 +32, # temp is originally in C; convert to F
                        elapsed_time_min=elapsed_time/60, # time is originally in seconds; convert to minutes
                        moving_time_min=moving_time/60,
                        average_speed_mpm=round(96.56064/average_speed,2), # speed is in km/hour, we want miles/min
                        max_speed_mpm=round(96.56064/max_speed,2),
                        total_elev_gain_ft=total_elevation_gain*3.281, # elevation is in meters, we want feet
                        elev_high_ft=elev_high*3.281,
                        elev_low_ft=elev_low*3.281
                        )
# split out minutes and hours into a more typical running pace format, for readability
runs <- runs %>% separate(average_speed_mpm, into=c("average_speed_mins", "average_speed_secs"), remove=FALSE)
runs <- runs %>% separate(max_speed_mpm, into=c("max_speed_mins", "max_speed_secs"), remove=FALSE)
runs <- runs %>% mutate(average_speed_mins=as.numeric(average_speed_mins),
                        max_speed_mins=as.numeric(max_speed_mins),
                        max_speed_secs=as.numeric(formatC(round(as.numeric(max_speed_secs)/100*60,2),width=2,format="d",flag="0")),
                        average_speed_secs=as.numeric(formatC(round(as.numeric(average_speed_secs)/100*60,2),width=2,format="d",flag="0")))
```

## Garmin Data
```{r, warning=FALSE, message=FALSE}
sleep <- read.csv("data/sleep/2021-06-23_2021-10-01_97013065_sleepData.csv")
sleep <- rbind(sleep, read.csv("data/sleep/2021-10-01_2022-01-09_97013065_sleepData.csv"))
sleep <- rbind(sleep, read.csv("data/sleep/2022-01-09_2022-04-19_97013065_sleepData.csv"))
sleep <- rbind(sleep, read.csv("data/sleep/2022-04-19_2022-07-28_97013065_sleepData.csv"))

drop <- names(sleep) %in% c("sleepWindowConfirmationType", "retro")
sleep <- sleep[!drop]
```

### Conversions
```{r}
sleep <- sleep %>% 
  mutate(totalSleepHours=(sleep$deepSleepSeconds + sleep$lightSleepSeconds + sleep$remSleepSeconds + sleep$unmeasurableSeconds)/60/60)
```

## Combine Garmin and Strava
```{r}
runs$calendarDate <- as.Date(runs$start_date) # get just the date (not time) of the run
sleep$calendarDate <- as.Date(sleep$calendarDate) # convert to Date datatype

# join sleep data to run dataframe
runs <- runs %>% left_join(sleep, by="calendarDate")
```

## Summary Stats
Let's get a basic understanding of the variables we'll be using in the models.
```{r, warning=FALSE}
# create nicer titles for each facet
var_names <- c(`average_heartrate` = c(paste("Heartrate\nMean: ", round(mean(runs$average_heartrate,na.rm=TRUE),2))),
               `distance_miles` = c(paste("Distance (Miles)\nMean: ", round(mean(runs$distance_miles,na.rm=TRUE),2))), 
               `suffer_score` = c(paste("Suffer Score\nMean: ", round(mean(runs$suffer_score,na.rm=TRUE),2))),
               `temp_f` = c(paste("Temperature (F)\nMean: ", round(mean(runs$temp_f,na.rm=TRUE),2))),
               `totalSleepHours` = c(paste("Previous Night's Sleep (Hours)\nMean: ", round(mean(runs$totalSleepHours,na.rm=TRUE),2))))

runs %>% 
  select(c("average_speed_mpm","distance_miles", "suffer_score", "average_heartrate", "temp_f", "totalSleepHours")) %>% 
  pivot_longer(!average_speed_mpm,names_to="Variable",values_to="Value") %>% 
  ggplot(aes(x=Value,fill=Variable)) + 
  geom_histogram() +
  scale_fill_manual(values=brewer.pal(n=5,name="Set1")) +
  facet_wrap(~Variable, scales="free", labeller = as_labeller(var_names)) +
  labs(title="Distribution of Each Model Factor", y="Count") +
  theme_minimal() +
  theme(legend.position="none")
```



## Models: Predicting Average Pace
### Strava Data Only
Let's start by investigating the effect of "within-run" factors on pace. The variables we'll use are distance, "suffer score", heartrate, and temperature.
```{r, warning=FALSE}
model_all <- lm(average_speed_mpm ~ distance_miles + suffer_score + average_heartrate + temp_f, data=runs)
summary(model_all)
qqnorm(residuals(model_all))

# find influential points
cooksd <- cooks.distance(model_all)
plot(cooks.distance(model_all))
influential <- as.numeric(names(cooksd)[(cooksd > 1)])
# look at this line in the data set
runs[influential,]

# remove the most influential point
runs <- runs[-influential,]

# re-run the model
model_all <- lm(average_speed_mpm ~ distance_miles + suffer_score + average_heartrate + temp_f, data=runs)
summary(model_all)
```

The important thing to keep in mind when interpreting the output of the model is that a higher number for pace means a slower pace (an 11 minute mile is slower than an 8 minute mile).

Distance: Distance is not a significant predictor in this model, with p = 0.946. It is weakly associated with pace in a negative direction - longer runs tend to have a lower (faster) pace, which is interesting and slightly unexpected.
Suffer score: Suffer score is not a significant predictor in this model, with p = 0.510. It is weakly associated with pace in a positive direction - higher suffer scores are associated with higher numbers for pace (slower runs).
Heartrate: Heartrate is a significant predictor of pace with p = 0.022. It is associated with pace in a negative direction - higher heartrates are associated with lower (faster) paces.
Temperature: Temperature is the strongest predictor in the model, with p < 0.001. As temperature increases, running pace also increases (i.e., when it is hotter, I tend to run slower).

```{r}
# plot residuals vs fitted values
plot(x=fitted.values(model_all), y=residuals(model_all))

# qq plot of residuals
qqnorm(residuals(model_all))

# plot residuals in order
plot(residuals(model_all))

```


Because temperature was such a strong predictor in the first model, I wanted to know what would happen if I removed it from a subsequent model - maybe its presence overshadowed some smaller but still meaningful effects of other variables.
```{r}
model_no_temp <- lm(average_speed_mpm ~ distance_miles + suffer_score + average_heartrate, data=runs)
summary(model_no_temp)
```

With temperature removed from the model, other variables became significant predictors.

Distance: Distance is a significant predictor in this model, with p = 0.038. It is associated with pace in a negative direction - longer distances are associated with lower (faster) paces.
Suffer score: Suffer score is also a significant predictor in this model, with p = 0.021. It is associated with pace in a positive direction - higher suffer scores are associated with higher numbers for pace (slower runs).
Heartreate: Heartrate is no longer a significant predictor of pace in this model, with p = 0.068. It is associated with pace in a negative direction - higher heartrates are associated with lower (faster) paces.

```{r}
# plot residuals vs fitted values
plot(x=fitted.values(model_no_temp), y=residuals(model_no_temp))

# qq plot of residuals
qqnorm(residuals(model_no_temp))

# plot residuals in order
plot(residuals(model_no_temp))
```




Now that we have two models, we need to compare them to see which model explains more variance in the data while using the fewest predictors.
```{r}
AIC(model_all,model_no_temp)
```

The model with temperature included has a lower AIC score, which means it does a better job of explaining our data, even though it has one more predictor. This is the model we will use going forward.


### Include Garmin Data
Now let's see if including biological data (sleep) adds any value.
```{r}
model_sleep <- lm(average_speed_mpm ~ distance_miles + suffer_score + average_heartrate + temp_f + totalSleepHours, data=runs)

summary(model_sleep)
```
Distance: Distance is not a significant predictor in this model, with p = 0.927. It is weakly associated with pace in a positive direction - longer runs tend to have a higher (slower) pace.
Suffer score: Suffer score is not a significant predictor in this model, with p = 0.639. It is weakly associated with pace in a positive direction - higher suffer scores are associated with higher numbers for pace (slower runs).
Heartrate: Heartrate is a significant predictor of pace with p = 0.029. It is associated with pace in a negative direction - higher heartrates are associated with lower (faster) paces.
Temperature: Temperature is the strongest predictor in the model, with p < 0.001. As temperature increases, running pace also increases (i.e., when it is hotter, I tend to run slower).
Sleep: Hours of sleep the night before a run is not a significant predictor, with p = 0.162. It is weakly associated with pace in a negative direction - getting more sleep is associated with lower (faster) pace.


```{r}
# plot residuals vs fitted values
plot(x=fitted.values(model_sleep), y=residuals(model_sleep))

# qq plot of residuals
qqnorm(residuals(model_sleep))

# plot residuals in order
plot(residuals(model_sleep))
```


Let's compare the model with sleep to the model without sleep.
```{r}
AIC(model_all,model_sleep)
```
The AIC scores are very similar. The model that includes sleep has a slightly lower score, which means it does a bit better job of explaining our data. This is the model we will visualize.

## Visualize the Model
```{r, warning=FALSE}
var_names <- c(`average_heartrate` = "Heartrate",
               `distance_miles` = "Distance (Miles)", 
               `suffer_score` = "Suffer Score",
               `temp_f` = "Temperature (F)",
               `totalSleepHours` = "Previous Night's Sleep (Hours)")


runs %>% 
  select(c("average_speed_mpm","distance_miles", "suffer_score", "average_heartrate", "temp_f", "totalSleepHours")) %>% 
  pivot_longer(!average_speed_mpm,names_to="Variable",values_to="Value") %>% 
  ggplot(aes(x=Value,y=average_speed_mpm,color=Variable)) + 
  geom_jitter(alpha=.5) +
  geom_smooth(aes(fill=Variable), alpha=.2, method=lm) +
  scale_color_manual(values=brewer.pal(n=5,name="Set1")) +
  scale_fill_manual(values=brewer.pal(n=5,name="Set1")) +
  facet_wrap(~Variable, scales="free_x", labeller = as_labeller(var_names)) +
  labs(title="Relationship of Each Model Factor with Running Pace", y="Pace (Miles per Minute)") +
  theme_minimal() +
  theme(legend.position="none")



```



## Other Investigations
### Group runs into seasons and look at mean pace across them
Temperature was a very strong predictor in the model, so let's take a closer look at that variable.

#### Data manipulation
To get a better idea of running pace variation across seasons, we will categorize each run by the season it occurred in.
```{r}
runs <- runs %>% mutate(month=as.integer(format(runs$calendarDate, "%m")))
winter <- c(12,1,2)
spring <- c(3,4,5)
summer <- c(6,7,8)
fall <- c(9,10,11)
runs <- runs %>% mutate(season=case_when(month %in% winter ~ "Winter",
                                 month %in% spring ~ "Spring",
                                 month %in% summer ~ "Summer",
                                 month %in% fall ~ "Fall"),
                        season=factor(season))
```

#### Plots
This plot shows the distribution of pace within each season. The mean pace for each season is shown by the black dot.
```{r}
runs %>% 
  group_by(season) %>% 
  summarize(mean_pace=mean(average_speed_mpm),
            mean_temp=mean(temp_f,na.rm=TRUE)) %>% 
  ggplot(aes(x=season,y=mean_pace)) +
  geom_bar(stat="identity") +
  labs(title="Mean Running Pace Across Seasons", x="Season", y="Pace") +
  theme_minimal() +
  theme(plot.title=element_text(hjust=0.5))

runs %>% ggplot(aes(x=average_speed_mpm)) +
  geom_dotplot(binwidth=.1,aes(fill=season)) 
  
runs %>% ggplot(aes(x=season,y=average_speed_mpm)) +
  geom_dotplot(binaxis = "y", stackdir = "center", binwidth=.1, aes(fill=season))


fall_color <- brewer.pal(n=9,name="PuOr")[2]
winter_color <- brewer.pal(n=9,name="RdGy")[7]
spring_color <- brewer.pal(n=9,name="RdYlGn")[8]
summer_color <- brewer.pal(n=9,name="Spectral")[6]


runs %>% ggplot(aes(x=season, y=average_speed_mpm)) +
  geom_violin(trim=FALSE, aes(fill=season)) +
  xlim(limits=c("Fall", "Winter", "Spring", "Summer")) +
  scale_fill_manual(values=c(fall_color,spring_color,summer_color,winter_color),guide="none") +
  stat_summary(fun="mean", geom="pointrange", color="black") +
  labs(title="Running Pace by Season",x="Season", y="Pace (Minutes per Mile)") +
  #geom_dotplot(binaxis='y', stackdir='center', dotsize=1, alpha=.3) +
  geom_jitter(shape=16, position=position_jitter(0.2), alpha=.3) +
  theme_minimal()

# gradient fill - https://stackoverflow.com/questions/41087157/gradient-fill-violin-plots-using-ggplot2

```

#### ANOVA
```{r}
season_model <- aov(runs$average_speed_mpm ~ runs$season)
anova(season_model)
qqnorm(residuals(season_model))
```

 